\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-05 k-planar.tex
more updates

@@

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.  We let $J:=G^\mathcal{P}/\mathcal{S}$ denote the resulting quotient graph and we let $V(J)\subseteq V(T)$ in the obvious way, so that each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\mathcal{P}}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $J$.)

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
		\begin{compactenum}
			\item No edge of $F$ is crossed by any edge of $G$; and
			\item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
			\begin{compactenum}
				\item $F[P_i]$ is a path; and
				\item $|V(P_i)\cap L_j| \le 15$ for every integer $j\ge 0$.
			\end{compactenum}
		\end{compactenum}
	\note{DW}{Should the vertices already grabbed by $P_i$ be mentioned in the setup? The new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 15$;
		\item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$
		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}; and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

		\item $G^+[\overline{Y}^+-F]$ is connected. To see this, first observe that $\overline{Y}-F$ is connected, and then observe that every vertex of $\overline{Y}^+$ is either a vertex of $\overline{Y}$ or adjacent to a vertex of $\overline{Y}$.
	\end{compactenum}
	Since $G^+$ is planar, the second point implies that $H=G^+/\mathcal{P}$ is planar.
	% \footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.} The first point then ensures that $G/\mathcal{P}$ is also planar.
\end{proof}

@@

Note that \citet{DJMMUW20} previously proved the bound of
$O(g^{k+2})$ using \cref{GenusProduct} and an ad-hoc method. Our result provides a better bound when $g>2^{k^3}$. In the case of 1-planar graphs we can improve further. \citet{ABGKP18} proved that every planar graph with treewidth at most $3$ has queue-number at most $5$. Thus the graph $H$ in \cref{1-planar} has queue-number at most $5$. \cref{qn,1-planar} then imply:

@@

\citet{PS21} and \citet{DFMS21} use \cref{ps} and \cref{PlanarProduct}(b), respectively, to show that $\chi_p(G)$ is polynomial in $p$ when $G$ is planar or of bounded Euler genus.  We use the following lemma due to \citet[Lemma~15]{pilipczuk.siebertz:polynomial-arxiv}.

\note{DW}{Is \cref{p-centered-treewidth} in the journal version \citep{PS21}?}

\citet{pilipczuk.siebertz:polynomial-soda} and \citet{debski.felsner.ea:improved} use \cref{ps} and \cref{PlanarProduct}(b), respectively, to show that $\chi_p(G)$ is polynomial in $p$ when $G$ is planar or of bounded Euler genus.  We use the following lemma due to \citet[Lemma~15]{pilipczuk.siebertz:polynomial-arxiv}.

\begin{lem}[\citep{pilipczuk.siebertz:polynomial-arxiv}]
\label{p-centered-treewidth}
  Every graph $H$ of treewidth at most $t$ has $\chi_p(H)\leq \binom{p+t}{t}$.
\end{lem}

\note{DW}{In the journal version \citet{DFMS21} (under the insistence of the referee :-) ) proved that $\chi_p(H_1\boxtimes H_2) \leq \chi_p(H_1) \cdot \chi(H_2^p)$ where $H_2^p$ is the $p$-th power of $H_2$. Apply this with $H_2:= P \boxtimes K_\ell$. Then $H_2^p= P^p\boxtimes K_\ell$ and $\chi(H_2^p)=(p+1)\ell$. So 
	$\chi_p(H\boxtimes P \boxtimes K_\ell) \leq \chi_p(H_1) \cdot (p+1)\ell$ }

\note{DW}{I suggest we replace the next paragraph by 	``\citet{DFMS21} proved that 
$\chi_p(H \boxtimes J) \leq \chi_p(H) \cdot \chi(J^p)$ for all graphs $H$ and $J$, which implies the following lemma'' and delete the proof. Okay?}

@@

\begin{proof}
	By \cref{PartitionProduct}, $G$ has an $H$-partition $(\mathcal{L}=\langle L_0,L_1,\ldots\rangle, \mathcal{P}=(B_x:x\in V(H))$ with layered width at most $\ell$. Use a product colouring $\phi:V(G)\to \{1,\ldots,\ell\}\times\{0,\ldots,p\}\times\{1,\ldots,\chi_p(H)\}$.  For each integer $i\ge 0$ and each $x\in V(H)$, assign the colour $\phi(v):=(\alpha(v),\beta(v),\gamma(v))$ to each vertex $v\in L_i\cap B_x$ such that:
   \begin{compactenum}
     \item $\alpha(v)$ is unique among $\{\phi(w): w\in L_i\cap B_x\}$, which is possible
     since $|L_i\cap B_x|\le \ell$,
     \item $\beta(v)= i\bmod (p+1)$, and
     \item $\gamma(v)=\gamma(x)$ where $\gamma:V(H)\to\{1,\ldots,\chi_p(H)\}$ is a $p$-centered colouring of $H$.
   \end{compactenum}
 To show this is a $p$-centered colouring, consider some connected subgraph $X\subseteq G$.

\referee{2}{Page 18. In point 1 in the proof of Lemma 9, I think '$\phi(w)$' should
be '$\alpha(w)$'.}

@@

\referee{2}{Page 20.  Where does the bound $d(d-3)/2$ come from?  The naive bound I
compute is $d(d-1)/2$.}

>>>>

\note{DW}{Use $\mathcal{S}$ for the given shortcut system and $\mathcal{P}$ for the partition?}

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.  Let $J:=G^\mathcal{P}/\mathcal{S}$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\mathcal{P}}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $J$.)

@@

@@

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
		\begin{compactenum}
			\item No edge of $F$ is crossed by any edge of $G$; and
			\item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, for some $k\in\{1,2,3\}$ such that for each $i\in\{1,\ldots,k\}$,
			\begin{compactenum}
				\item $F[P_i]$ is a path; and
				\item $|V(P_i)\cap L_j| \le 15$ for every integer $j\ge 0$.
			\end{compactenum}
		\end{compactenum}
	\note{DW}{Should the vertices already grabbed by $P_1,P_2,P_3$ be mentioned in the setup? The new paths $Q_1,Q_2,Q_3$ need to avoid the vertices already grabbed by $P_1,P_2,P_3$. Basically, I am concerned about what happens when distinct vertical paths in $T$ want to grab the same vertex.}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that:
	\begin{compactenum}
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 15$;
		\item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$
		\note{DW}{Should this be $P_i\subseteq S_{x_i}$? (since the path grabs other vertices into its part)}; and
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

		\item $G^+[\overline{Y}^+-F]$ is connected. To see this, first observe that $\overline{Y}-F$ is connected, and then observe that every vertex of $\overline{Y}^+$ is either a vertex of $\overline{Y}$ or adjacent to a vertex of $\overline{Y}$.
	\end{compactenum}
	Since $G^+$ is planar, the second point implies that $H=G^+/\mathcal{P}$ is planar.
	% \footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.} The first point then ensures that $G/\mathcal{P}$ is also planar.
	\note{DW}{We need a sentence about why $H$ Has the claimed tree-decomposition.}
\end{proof}

@@

Note that \citet{DJMMUW20} previously proved the bound of
$O(g^{k+2})$ using \cref{GenusProduct} and an ad-hoc method. Our result provides a better bound when $g>2^{k^3}$. In the case of 1-planar graphs we can improve further. \citet{ABGKP20} proved that every planar graph with treewidth at most $3$ has queue-number at most $5$. Thus the graph $H$ in \cref{1-planar} has queue-number at most $5$. \cref{qn,1-planar} then imply:

@@

\citet{PS21} and \citet{DFMS21} use \cref{ps} and \cref{PlanarProduct}(b), respectively, to show that $\chi_p(G)$ is polynomial in $p$ when $G$ is planar or of bounded Euler genus.  We use the following lemma due to \citet[Lemma~15]{PS21}.

\begin{lem}[\citep{PS21}]
\label{p-centered-treewidth}
  Every graph $H$ of treewidth at most $t$ has $\chi_p(H)\leq \binom{p+t}{t}$.
\end{lem}

\note{DW}{In the journal version \citet{DFMS21} (under the insistence of the referee :-) ) proved that $\chi_p(H_1\boxtimes H_2) \leq \chi_p(H_1) \cdot \chi(H_2^p)$ where $H_2^p$ is the $p$-th power of $H_2$. Apply this with $H_2:= P \boxtimes K_\ell$. Then $H_2^p= P^p\boxtimes K_\ell$ and $\chi(H_2^p)=(p+1)\ell$. So 
	$\chi_p(H\boxtimes P \boxtimes K_\ell) \leq \chi_p(H_1) \cdot (p+1)\ell$. So I suggest we replace the next paragraph by 	``\citet{DFMS21} proved that 
$\chi_p(H \boxtimes J) \leq \chi_p(H) \cdot \chi(J^p)$ for all graphs $H$ and $J$, which implies the following lemma'' and delete the proof. Okay?}

@@

\note{DW}{For consistency with \cref{qn}, I suggest we combine \cref{p-centered,p-centered-treewidth}, and just write, ``For every graph $H$ of treewidth at most $t$ and for every path $P$, if $G\subseteq H\boxtimes P \boxtimes K_\ell$ then 
	\[\chi_p(G)\le \ell (p+1)\, \chi_p(H) \leq (p+1)\ell\, \tbinom{p+t}{t}.\]}

\begin{proof}
	By \cref{PartitionProduct}, $G$ has an $H$-partition $(\mathcal{L}=\langle L_0,L_1,\ldots\rangle, \mathcal{P}=(B_x:x\in V(H))$ with layered width at most $\ell$. Use a product colouring $\phi:V(G)\to \{1,\ldots,\ell\}\times\{0,\ldots,p\}\times\{1,\ldots,\chi_p(H)\}$.  For each integer $i\ge 0$ and each $x\in V(H)$, assign the colour $\phi(v):=(\alpha(v),\beta(v),\gamma(v))$ to each vertex $v\in L_i\cap B_x$ such that:
   \begin{compactenum}
     \item $\alpha(v)$ is unique among $\{\alpha(w): w\in L_i\cap B_x\}$, which is possible
     since $|L_i\cap B_x|\le \ell$,
     \item $\beta(v)= i\bmod (p+1)$, and
     \item $\gamma(v)=\gamma(x)$ where $\gamma:V(H)\to\{1,\ldots,\chi_p(H)\}$ is a $p$-centered colouring of $H$.
   \end{compactenum}
 To show this is a $p$-centered colouring, consider some connected subgraph $X\subseteq G$.

@@

\referee{2}{Page 20.  Where does the bound $d(d-3)/2$ come from?  The naive bound I
compute is $d(d-1)/2$.}
\note{DW}{consecutive neighbours don't have to be counted. I will revise}
